---
title: "Peer-graded Assignment: Getting and Cleaning Data Course Project"
output:
  html_document:
    df_print: paged
---

# hello there!

The following file holds all the procedures to process the given data. This version is Rstudio based, but an HTML is also provided for visualization.


# the assingment instructions

  A full description is available at the site where the data was obtained:
  
  http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones
  
  Here are the data for the project:
  
  https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip
  
  You should create one R script called run_analysis.R that does the following.
  
      1. Merges the training and the test sets to create one data set.
      2. Extracts only the measurements on the mean and standard deviation for each measurement.
      3. Uses descriptive activity names to name the activities in the data set
      4. Appropriately labels the data set with descriptive variable names.
      5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
  
  Good luck!


# 1 Merges the training and the test sets to create one data set.

## Getting data

All data should be download and decompressed into the standard directory within the file. The variable below navigates from getwd() into it. Set your setwd() accordilly.
In fact, the best approach is to be in the same directory as "UCI HAR Dataset".

```{r}
getwd()
root <- "UCI HAR Dataset"
```


## Loading data

The column names are in "features.txt" while the remaining two columns (subject and activity) are nameless.

I want a vector with column names:

```{r}
# move into the directory
file <- file.path(root, "features.txt")
column.names <- read.table(file) # the file
                           # header = FALSE, # there is no header
                           # sep = " ", # white space rulez
                           # strip.white = TRUE) # but not so much

# i am also getting rid of the numbers by integrating
# them into the names

column.names <- with(column.names, paste(V1, V2))

# adding names for subject and activity
# i am also adding a new column to track
# the source of each row from the original
# datasets

column.names <- c("subject", "activity", as.character(column.names), "source")
```

## The data

Now, lets load the data. train set first.

Data is stored this way:

  * dir (test/train)
  * subject_?.txt (the test subjects)
  * X_*.txt (the data)
  * Y_*.txt (the activities)
  
```{r}
## loading the test subjects
file <- file.path(root, "train/subject_train.txt")
subjects <- read.table(file)

## activities
file <- file.path(root, "train/y_train.txt")
activities <- read.table(file)

## data
# loading this data may be tricky, the file loads better
# with read.table. check if the number of obs fits the
# remaing variables (7352 obs x 561 vars)
file <- file.path(root, "train/X_train.txt")
data <- read.table(file)

head(subjects)
head(activities)
head(data)
```


## Assembling

With all main data available, the main data.frame may be assembled.

```{r}
# data.frame itself the handle multiple inputs,
# there is no need for rbind/cbind now
# factor() populates the last column: source
dataset <- data.frame(subjects, activities, data, factor("train"))

# put columns.names on it
names(dataset) <- column.names

head(dataset)
```

And done. I will keep the names like this for the following reasons:
  * the numbers help me out locating the columns, this is useful for graphs
  * some parts of the names are similar and useful for partial matching

Now we reapeat the load process for the test set and merge them togheter.

## Loading data, 2

lets reuse the code from before.

```{r}
## loading the test subjects
file <- file.path(root, "test/subject_test.txt")
subjects <- read.table(file)

## activities
file <- file.path(root, "test/y_test.txt")
activities <- read.table(file)

## data
# loading this data may be tricky, the file loads better
# with read.table. check if the number of obs fits the
# remaing variables (2947 obs x 561 vars)
file <- file.path(root, "test/X_test.txt")
data <- read.table(file)

head(subjects)
head(activities)
head(data)
```

## Assembling, 2

I do prefer to create a second dataset before merge.

```{r}
# a second dataset, with more data
dataset2 <- data.frame(subjects, activities, data, factor("test"))

# same strucutre as before
names(dataset2) <- column.names

head(dataset2)
```

## Merging

Since the datasets where created with same columns, they can be merged with rbind.

```{r}
# since they are same sized and
# with same column names
# rbind() behaves well
data <- rbind(dataset, dataset2)
```

And done again. Now all data is in "data" variable. Let's save the first output.

```{r}
# the following code clears the memory from duplications.
# it is disabled by default

#rm(activities, dataset, dataset2, subjects, tmp)

# save data
#save(data, file = "week4.project.csv")
write.table(data, file = "week4.project.txt", row.names = FALSE)
```


# 2. Extracts only the measurements on the mean and standard deviation for each measurement.

## The selected columns

There are two ways of doing this:
  * Select manually the columns using its regular interval. This works
  because every 40 columns mean() and sd() reccurs.
  * Use grep.
  
```{r}
# the following regex selects only the required columns.
# the \\ are scape characters for ()
target <- grep("mean\\(\\)|std\\(\\)", column.names)

# i will also keep the first two

target <- c(1:2, target)

# with this I can easily restrict the data
selected <- data[, target]
```

# 3. Uses descriptive activity names to name the activities in the data set

## using factors

The names for the activities are in activity_labels.txt. One just need to substitute them in the data$activities factor.

```{r}
# load the data from file
file <- file.path(root, "activity_labels.txt")
labels <- read.table(file)

# do the substitution using a anonymous function
selected$activity <- (with(data, sapply(activity,
                       function(item) {labels$V2[item]})
          ))

selected$activity <- factor(selected$activity)
```

# 4 Appropriately labels the data set with descriptive variable names.

I like the names choosen by the authors, but some improvement can be done. $subject and $activity are set already, let us focus on the others.

The names follow the structure:

  * number
  * t/f for time or frequency domain
  * the true variable name
  * the conversion (mean, std...)
  * the axis

There are some exceptions that will be dealt with on the way.

I have no need for numbers anymore, I want to use dots to separate fields like we do in R, remove the function calls from function names (so I can use them directly in R) and move the axis first. A regex is needed:

```{r}
## pattern 1
# check names
names(selected)

# i am using perl like regex. scapes need to be double scaped
# grouping is per parenthesis
regex <- "([0-9]+ )(t|f)([A-Za-z]+)\\-(mean|std)\\(\\)\\-(X|Y|Z)"
  #groups:  1       2     3             4           -       5

# check regex
grep(regex, names(selected))

#change names
names(selected) <- gsub(regex,"\\2.\\5.\\4.\\3", names(selected))

## pattern 2
names(selected)
regex <- "([0-9]+ )(t|f)([A-Za-z]+)\\-(mean|std)\\(\\)"
# groups  1         2   3               4
grep(regex, names(selected))
names(selected) <- gsub(regex,"\\2.\\4.\\3", names(selected))

# all lower case
names(selected) <- tolower(names(selected))
```

# 5 From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

split + lapply + sapply + data.frame produce a useful result. but aggregate() is better.

the data will be like this: each line has the mean of the selected variables for each subject x activity combination.

```{r}
# to ensure a good behaviour
selected$subject <- factor(selected$subject)

factor(selected$subject)

# create a table of averages
averages <- aggregate(selected, list(selected$subject, selected$activity), mean)

# rename and remove some columns
averages$subject <- averages$Group.1
averages$activity <- averages$Group.2
averages <- averages[, 3:70]

# save data
#save(averages, file = "averages.R")
write.table(averages, file = "averages.csv", row.names = FALSE)
```

